---
title: 'Class Project in Two Parts: Central Limit Theorem & Inferential Data Analysis'
author: "Paul Clark"
date: "February 5, 2017"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 5
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 5
---
---

# Central Limit Theorem (Simulation)

## Overview

We investigate the *exponential* distribution in R and compare it with the Central Limit Theorem. The distribution is simulated via `rexp(n, lambda)`, and both mean and standard deviation are `1/lambda`. We set `lambda = 0.2` and investigate the distribution of 1000 samples of 40 exponentials each.

**Review criteria**  

  1. Did you show where the distribution of means is centered and compare it to the theoretical center of the population distribution?
  2. Did you show how variable it is and compare it to the theoretical variance of the distribution?

## Simulations
We will show the properties of the distribution of the mean and variance of samples of 40 exponentials.  

First, we simulate the data:
```{r exp_simulation}
# We simulate S = 1000 samples of N = 40 exponentials each, all with lambda = 0.2,
# storing their means and variances in a data frame
S <- 1000
N <- 40
lambda <- 0.2
sim_data <- matrix(rexp(n = S*N, rate = lambda), S, N)
sim_mean <- apply(sim_data, 1, mean)
sim_var <-  apply(sim_data, 1, function(x) {var(x)*N/(N-1)}) # sample variance uses N-1
```

### Sample Mean vs. Theoretical Mean
Next, we show our simulated distribution of sample means and compare it to the theoretical mean of the population.
```{r means_histogram}
# A histogram of sample means, with vertical line showing position of theoretical mean
# within the histogram


```
### Sample Variance vs. Theoretical Variance
We show the sample variance and compare it to the theoretical variance of the population.
```{r variance_histogram}
# Following slide XX in lecture YY, a histogram of sample variances, with
# vertical line showing position of theoretical variance within the histogram

```
### Normality
We now show that the distribution is approximately normal.
```{r means_and_pop_histograms}
# A two-facet plot contrasting the randomly sampled distribution of means with
# the randomly sampled distribution of the population

```
By visual inspection, the population distibution is clearly not normal. Therefore, we focus our evaluation on the distribution of means.
#### Mean vs. Median
The Mean and Median of a normal distribution should be equal. This is approximately true for the distribution of means.
```{r mean_vs_median}

```
#### Percentage of distribution within 1 to 3 standard deviations of the mean

```{r percent_of_distribution}

```

Sample Mean versus Theoretical Mean: Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.  

Sample Variance versus Theoretical Variance: Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.  

Distribution: Via figures and text, explain how one can tell the distribution is approximately normal.  

# Basic Inferential Data Analysis 

**Review Criteria**

  3. Did you perform an exploratory data analysis of at least a single plot or table highlighting basic features of the data?
  4. Did the student perform some relevant confidence intervals and/or tests?
  5. Were the results of the tests and/or intervals interpreted in the context of the problem correctly?
  6. Did the student describe the assumptions needed for their conclusions?

Now in the second portion of the project, we're going to analyze the ToothGrowth data in the R datasets package.  

Load the ToothGrowth data and perform some basic exploratory data analyses  

Provide a basic summary of the data.  

Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)  

State your conclusions and the assumptions needed for your conclusions.